Enhancing Accelerometer-based Activity Recognition with Capacitive Proximity Sensing

### 2012-06-22 Paper/Paper.tex
prefinal version of the paper

\maketitle
\begin{abstract}
Activity recognition with an accelerometer is a common investigated research topic and enables the detection of basic activities like sitting, walking or standing. Recent work in this area adds different sensing modalities to the inertial data to collect more information of the users environment and to boost activity recognition. This work presents a sensor prototype which consists of an accelerometer and a capacitive proximity sensor that senses the users activity based on the combined sensor values. We show that our proposed approach improves the recognition rate significantly for detecting activities of daily living (e.g. sitting, preparing dinner, eating, drinking, etc.) by adding capacitive proximity sensor data to the inertial data.
\keywords{activity recognition, capacitive proximity sensors, ambient assisted living, user context}
\end{abstract}

@@

Persons affected from physical and mental restrictions and their care givers can profit significantly from unobtrusive activity monitoring solutions. For example, formal care givers could equip a person with a wearable activity monitoring solution that evaluates the course of a disease or the influence of an adapted medication. This scenario is especially relevant for persons suffering from dementia who have limitations in organizing their daily activities. A simple wearable activity monitoring solution could analyze activities performed in daily life, such as drinking, eating and sleeping habits.

@@

Capacitive proximity sensors on the other hand can indirectly measure the distance and nature of a grounded object within reach. This means that the measurement result depends on the object's distance, its size and the material it is made of. In this work, we show that an accelerometer and a capacitive proximity sensor can be used to improve activity recognition in activities of daily living.
% For this we obtained a wrist-worn activity data logger\footnote{HedgeHog Activity Logger -- \url{http://www.ess.tu-darmstadt.de/hedgehog}} and integrate a capacitive proximity sensor unobtrusively into the wristband.
We obtained an open-hardware and open-source wrist-worn activity data logger \cite{hedgehog} and integrated a capacitive proximity sensor invisibly into the wristband.

@@

% This paper is structured as follows: We give a short insight in related work in chapter 2, followed by describing the hardware set-up used for this study in chapter 3. The experiment and evaluation will be described in chapter 4, showing how data has been obtained and how activity recognition is being improved. This work is concluded in chapter 5 with an outlook to future work.

@@

Activity recognition research relying on wearable sensors mostly considers inertial data from the participants body to infer performed activities, such as in the works of \cite{Ravi2005,Bao2004,Srinivasan2010,Amft2005}. The acceleration data is often augmented with data from sensors such as gyroscopes \cite{Holleczek_2010}, magnetometers \cite{Altun_2010}, ambient light \cite{Borazio2012} or ambient and skin temperature \cite{Krause_2003}, aiming to extract a more detailed environmental user context. Researchers in \cite{wyss2010recognition} use heart rate information as an addition to the accelerometer data to detect military related activities like lifting and lowering loads or even digging. In \cite{Ward_2006}, workshop assembly activities are detected by augmenting acceleration sensors with microphones.

The works of \cite{Fishkin_2005} and \cite{patterson2005fga} show that detecting touched and used objects can be very helpful for activity recognition. By using RFID readers that are embedded in gloves or bracelets at the wrist and RFID tags attached to various objects of interest, one can detect the object grasped and used by the user, thus aiding the activity recognition in various application scenarios, such as activities of daily living \cite{phealth:maja} and \cite{Philipose_2004}, activity tracking in car manufacturing \cite{Stiefmeier08}, or household and gardening activities \cite{berlin_laerhoven_tei_2010}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

This section presents the two components of the required hardware, the wrist-worn activity data logger tailored to capture acceleration data, and the capacitive proximity sensor used for distance measurements.

@@

The sensor node itself is built around the low-power Microchip microcontroller (PIC18F46J50) featuring an accelerometer sensor (ADXL345) to capture human motion, light and ambient temperature sensors and a microSD flash card for locally storing the sensor data. The sensor is powered by a $200mAh$ lithium polymer battery, which allows for two weeks of continuous recording on a single battery charge. A USB port is used to configure the sensor (e.g. setting the sensitivity of the accelerometer), to access the stored sensor data, and to recharge the battery. A plastic case nicely packages and protects the sensor to be worn at the wrist (as shown in Fig. \ref{fig:sensornode}).

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{Images/hardware_sensor_2.jpg}
	\caption{The inertial data logger featuring a low-power microcontroller, a 3 axis accelerometer, a microSD flash card for storing the sensor data and a USB connector for accessing the stored data (on the right) is being powered by a small lithium polymer battery and is packaged into a plastic case to be worn at the wrist (a version with a OLED display on the left).}
	\label{fig:sensornode}
\end{figure}

The 3D accelerometer sensor is being sampled at $100Hz$, resulting in 10ms equidistant measurements. For efficiency reasons, the sensor data is run-length encoded before being stored locally to the microSD card. The HedgeHog can be extended with further sensors tailoring different application scenarios. For our scenario, we have added a capacitive proximity sensor that is described in detail in the next section.

@@

The capacitive proximity sensor performs measurements in loading mode. Two electrodes are integrated into the wristband, one sensing electrode and one shielding electrode that aims to eliminate the influence of the underlying skin. Without shielding no meaningful measurements could be performed, since the arm's skin is a very proximate grounded object to the sensing electrode. The sensor draws a supply current of 1mA at 3.3V when active and qualifies it for wearable proximity sensing applications. In the following a virtual capacitor denotes the capacitance between the sensing electrode and the environment. The capacitance of the sensing electrode to environmental objects increases with closer distances.

@@

\begin{figure}
	\centering
		\includegraphics[width=1.00\textwidth]{Images/capacitive_sensor_wristband_2.jpg}
	\caption{The hardware prototype at a glance: HedgeHog activity logger at the lower right, the capacitive sensor unit at the lower left, and the wristband with the sensing and the shield electrodes on-top each other covered with adhesive tape for isolation purposes.}
	\label{fig:cap_sensor}
\end{figure}

@@

The opening door activity has very poor recognition rates without the data from the proximity sensor. The average f-measure could be increased from 35.5\% to 46.2\%. A plot of the activity is given in Figure \ref{fig:opendoor}. The capacitive proximity sensor shows two approaches to the door knob, one for opening the door (2s) and one for closing the door (9s). The acceleration sensor captures relevant data in the time in which the person moves into the room and the hand changes from the outer to the inner door knob (5 - 7s). The recorded data for this activity shows strong correlations between all test persons. The confusion matrices show that the ``open door'' activity was often confused with the ``sitting'' activity, because of the horizontal orientation of the hand. Due to the usage of the capacitive sensor, the recall for that class and confusion with the sitting activity improved. [wording, orientation vs little motion ?]

@@

After closing the door, the participants were supposed to sit down on the couch. It turned out that there are great variations of the sitting posture and the corresponding hand positions. Many users tapped with their fingers or hands while sitting, changed their sitting positions very frequently, or were even talking and gesticulating, as shown in Figure \ref{fig:sitting}. In this case, it is obvious that the data from the acceleration sensor is very difficult to interpret as there are numerous changes in the axial orientation of the sensor. However, the capacitive proximity sensor is able to indicate when a hand is placed on the surface of the couch. Especially for this particular participant, the f-measure increased from 50.3\% to 60.4\%, while the average f-measure improved from 68.0\% to 74.4\%.

In the following, the participant were instructed to lie  down on the couch. Again, there were great variations in how this activity was performed by the participants. For example, some of them crossed their hands under their head, or placed them on their stomach. For this class, the average f-measure could only be increased by 2.4\%, from 81.2 to 83.6\%. Considering some test persons, this class was confused with the ``making bread'' activity and could be improved by the proximity sensor. [ wording! ]

@@

After that, the participants were asked to walk over to the kitchen and to put food and dishes from a shelf and a locker on the table. This activity involved direct interactions with various objects as well as proximity to furniture in the room (see Table \ref{tab:activities}). The capacitive proximity sensor was able to capture the proximity to the shelf and to the table. The average f-measure for this activity is rather low, but improved by 6.8\% from 53.8\% to 60.6\%. The worst performing participants for this activity reached an f-measure of 46.8\% without and 53.5\% with the capacitive sensor, while the best performing one reached 62.0\% and 64.5\% respectively. The quite low performance figures result from confusions with other activities, with a higher tendency to the ``prepare bread'' activity across all participants, most likely because there are also various objects involved. The capacitive sensor modality has a positive impact reducing the confusion with other activities.

\begin{figure}
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_5.pdf}
	\caption{An example of the ``preparing bread'' activity, where the participants had to put marmalade on a toast. The proximity sensor indicates the closeness to the table, while the acceleration sensor shows reoccurring movements.}
	\label{fig:prep_bread}
\end{figure}

Figure \ref{fig:prep_bread} shows an example instance of preparing a bread with marmalade. It is notable that the acceleration data does not seem to provide any characteristic patterns, while the proximity sensor indicates a table, plate, or other objects in immediate distance. This activity showed a high improvement in the average f-measure by 10\%, from 49.0\% to 59.8\%, where the data delivered by the capacitive proximity sensor is taken into account. The making bread class was often confused with the sitting class for some users, probably due to lots of motion during the sitting, as mentioned previously. For other users, making bread was confused with eating or drinking. Using the new input modality, confusion across users could be reduced.

\begin{figure}
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_6.pdf}
	\caption{An eating person who took 5 bites from the bread. After each bite, the hand is placed on the table.}
	\label{fig:eating}
\end{figure}

When considering the eating activity, the impact of the new capacitive proximity sensor on the classification performance is quite low, as the chosen features are able to distinguish it from other activities. Some of the participants ate their bread leaving their hand close to the mouth, while others moved their hand up and down putting their bread aside on the plate, which also results in the performance range from 71.2 to 88.1\% without and 77.5 to 90.6\% with the proximity data. A good example of an eating activity is shown in Figure \ref{fig:eating}, where the participant took a few bites from the bread while putting it down every time. The average f-measure for that activity increased slightly by 2.7\% (from 79.5\% to 82.2\%).

\begin{figure}
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_7.pdf}
	\caption{An example of the ``drinking'' activity. The participant first pours some water into the glass and then takes three drinks of water. After each sip, he returns his arm to the table which can be observed in the characteristic patterns of the proximity sensor.}
	\label{fig:drinking}
\end{figure}

@@

Regarding the walking activity one can identify periodic changes in the acceleration as well as in the measured capacitance, illustrated in Figure \ref{fig:walking}. While walking, the capacitance between the wristband and the leg increases when the wristband is located close to the body and decrease when the wristband moves away. There were problems in distinguishing this activity from get things that could be bettered by using the new input modality. The classification improvement for this activity accounts to 12.2\% boosting the average f-measure from 41.7\% without to 53.9\% with the new sensor. Due to the low performance results and the characteristic periodic signal shape, it would help to consider frequency domain features, as it is often applied in related work.

\begin{figure}
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/eugen_9.pdf}
	\caption{During the sleeping activity the data from both sensors remains constant. One draw conclusions about the coverage of the arm with either couchons, blankets or the proximity to the mattress.}
	\label{fig:sleeping}
\end{figure}

The sleeping activity (an example shown in Figure \ref{fig:sleeping}) was classified with an average f-measure of 83.2\%, which increased to 86.9\% when using the proximity data. In this case, the capacitive proximity sensor is able to capture the surrounding cushions and blankets, as well as the body or head of the participant. The accelerometer data and the proximity data has larger periods with a constant signal. [further details?]

\begin{figure}
	\setlength{\tabcolsep}{2.4pt}
	\centering
	\begin{scriptsize}
	\begin{tabular}{|ccccccccc|p{0.01cm}|ccccccccc|p{0.01cm}|l}
% 		\toprule
		\multicolumn{9}{c}{without proximity data} &\multicolumn{1}{c}{}& 
			\multicolumn{9}{c}{with proximity data} &\multicolumn{1}{c}{}&  \\ \cmidrule{1-9} \cmidrule{11-19}

		\multicolumn{1}{c}{a} & b & c & d & e & f & g & h & \multicolumn{1}{c}{i} &\multicolumn{1}{c}{}& 
			\multicolumn{1}{c}{a} & b & c & d & e & f & g & h & \multicolumn{1}{c}{i} 	&\multicolumn{1}{c}{}& $\leftarrow$ classified as\\
		\cline{1-9} \cline{11-19} \cline{21-21}
% 		\textbf{9} & 5 & 0 & 0 & 1 & 0 & 0 & 0 & 0 		&& \textbf{10} & 3 & 0 & 2 & 0 & 0 & 0 & 0 & 0 		&& a = open door	\\ 
% 		1 & \textbf{248} & 5 & 6 & 4 & 2 & 5 & 0 & 0 	&& 2 & \textbf{259} & 4 & 2 & 4 & 2 & 4 & 0 & 0 	&& b = sitting		\\ 
% 		0 & 3 & \textbf{212} & 3 & 2 & 1 & 6 & 0 & 8 	&& 0 & 5 & \textbf{213} & 2 & 2 & 1 & 8 & 0 & 4 	&& c = lying		\\ 
% 		0 & 8 & 2 & \textbf{44} & 7 & 8 & 1 & 1 & 3 	&& 3 & 5 & 2 & \textbf{52} & 9 & 3 & 1 & 1 & 1 		&& d = get things	\\ 
% 		1 & 7 & 1 & 11 & \textbf{37} & 9 & 2 & 0 & 10 	&& 1 & 9 & 3 & 13 & \textbf{46} & 6 & 3 & 0 & 0 	&& e = make bread	\\ 
% 		0 & 3 & 2 & 3 & 8 & \textbf{58} & 5 & 0 & 1 	&& 0 & 5 & 1 & 1 & 5 & \textbf{62} & 5 & 0 & 1 		&& f = eating		\\ 
% 		0 & 8 & 3 & 0 & 6 & 4 & \textbf{67} & 0 & 1 	&& 0 & 8 & 4 & 2 & 1 & 5 & \textbf{68} & 0 & 0 		&& g = drinking		\\
% 		1 & 0 & 0 & 1 & 0 & 0 & 0 & \textbf{12} & 0 	&& 1 & 0 & 0 & 1 & 0 & 0 & 0 & \textbf{13} & 0 		&& h = walking		\\
% 		0 & 1 & 2 & 10 & 6 & 1 & 1 & 0 & \textbf{312} 	&& 1 & 0 & 0 & 7 & 0 & 0 & 0 & 0 & \textbf{325} 	&& i = sleeping		\\ 
% 	
%		\textbf{11} & 0 & 5 & 4 & 1 & 0 & 0 & 0 & 0 	&& \textbf{17} & 0 & 0 & 2 & 0 & 0 & 0 & 0 & 0		&& a = open door    \\
%		0 & \textbf{136} & 8 & 1 & 11 & 0 & 0 & 0 & 0	&& 0 & \textbf{140} & 7 & 2 & 3 & 0 & 0 & 0 & 0		&& b = sitting      \\
%		4 & 20 & \textbf{151} & 1 & 7 & 2 & 1 & 1 & 0	&& 0 & 23 & \textbf{149} & 5 & 7 & 3 & 2 & 1 & 0	&& c = lying        \\
%		3 & 6 & 5 & \textbf{116} & 21 & 9 & 4 & 4 & 2	&& 1 & 1 & 9 & \textbf{129} & 25 & 4 & 2 & 4 & 1	&& d = get things   \\
%		0 & 5 & 2 & 5 & \textbf{81} & 11 & 2 & 0 & 0	&& 0 & 3 & 1 & 15 & \textbf{83} & 6 & 1 & 0 & 1		&& e = make bread   \\
%		0 & 0 & 0 & 0 & 9 & \textbf{167} & 19 & 0 & 3	&& 0 & 0 & 0 & 4 & 8 & \textbf{158} & 26 & 0 & 3	&& f = eating       \\
%		1 & 2 & 3 & 4 & 8 & 26 & \textbf{89} & 0 & 5	&& 0 & 0 & 8 & 5 & 5 & 29 & \textbf{92} & 1 & 4		&& g = drinking     \\
%		0 & 0 & 1 & 17 & 3 & 0 & 0 & \textbf{2} & 0		&& 0 & 1 & 2 & 10 & 3 & 0 & 0 & \textbf{9} & 0		&& h = walking      \\
%		0 & 0 & 0 & 2 & 0 & 8 & 2 & 0 & \textbf{78}		&& 0 & 0 & 0 & 2 & 0 & 6 & 1 & 0 & \textbf{81}		&& i = sleeping     \\

		\textbf{8} & 0 & 0 & 0 & 0 & 0 & 0 & 2 & 1				&& \textbf{11} & 0 & 0 & 2 & 1 & 0 & 0 & 0 & 1 		&& a = open door	\\ 
		0 & \textbf{140} & 0 & 5 & 2 & 0 & 0 & 0 & 0  		&& 0 & \textbf{140} & 0 & 5 & 2 & 0 & 0 & 0 & 0 	&&  b = sitting		\\ 
		0 & 0 & \textbf{294} & 1 & 9 & 10 & 1 & 1 & 0 	 	&& 0 & 0 & \textbf{309} & 1 & 0 & 4 & 0 & 0 & 1 	&& c = lying		\\ 
		1 & 7 & 1 & \textbf{50} & 17 & 9 & 1 & 4 & 1	 		&& 2 & 6 & 0 & \textbf{67} & 14 & 9 & 1 & 2 & 1 	&& d = get things	\\ 
		0 & 2 & 7 & 7 & \textbf{102} & 42 & 5 & 0 & 2			&& 1 & 1 & 0 & 13 & \textbf{120} & 24 & 7 & 1 & 1 && e = make bread	\\ 
		0 & 0 & 3 & 5 & 21 & \textbf{288} & 4 & 0 & 0 	 	&& 0 & 0 & 2 & 1 & 19 & \textbf{300} & 3 & 0 & 0 	&& f = eating		\\ 
		0 & 5 & 11 & 1 & 29 & 40 & \textbf{18} & 0 & 0	 	&& 0 & 4 & 4 & 0 & 26 & 57 & \textbf{16} & 0 & 0	&& g = drinking		\\
		2 & 0 & 0 & 2 & 0 & 0 & 0 & \textbf{11} & 0	 			&& 2 & 0 & 0 & 0 & 0 & 0 & 0 & \textbf{14} & 0 		&& h = walking		\\
		0 & 2 & 2 & 0 & 2 & 0 & 0 & 0 & \textbf{271}			&& 0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 & \textbf{276} 		&& i = sleeping		\\

		\cline{1-9} \cline{11-19} 
% 		\bottomrule
	\end{tabular}
	\end{scriptsize}

	\caption{Activity recognition evaluation revealing the positive impact of the capacitive proximity sensor. Here, we are comparing SVM classification presented as confusion matrices for an exemplary user, without the proximity data on the left, and with the proximity data on the right. Note: reject class is not included in the confusion matrix.}

	\label{fig:ConfusionMatrixForAnExemplaryUser}
\end{figure}

Figure \ref{fig:ConfusionMatrixForAnExemplaryUser} depicts an example for the SVM classification results for one of the participants.

@@

We have shown that current wearable inertial sensing devices for activity recognition can be significantly enhanced by integrating a capacitive proximity sensor into the wristband. Regarding the classification accuracy, we obtained an overall increase in correctly classified instances of TODO percent. Some classes, for instance ``preparing bread'' and ``drinking'', could benefit very much from proximity-related sensor data. For such classes the classification performance could be boosted by up to TODO percent.

>>>>

\maketitle
\begin{abstract}
Activity recognition with an accelerometer is a common investigated research topic and enables the detection of basic activities like sitting, walking or standing. Recent work in this area adds different sensing modalities to the inertial data to collect more information of the users environment and to boost activity recognition. This work presents a sensor prototype which consists of an accelerometer and a capacitive proximity sensor that senses the users activity based on the combined sensor values. We show that our proposed approach significantly improves the recognition rate for detecting activities of daily living (e.g. sitting, preparing dinner, eating, drinking, etc.) by adding capacitive proximity sensor data to the inertial data.
\keywords{activity recognition, capacitive proximity sensors, ambient assisted living, user context}
\end{abstract}

@@

Persons affected from physical and mental restrictions and their care givers can profit significantly from unobtrusive activity monitoring solutions. For example, formal care givers could equip a person with a wearable activity monitoring solution that evaluates the course of a disease or the influence of an adapted medication. This scenario is especially relevant for people suffering from dementia who have limitations in organizing their daily activities. A simple wearable activity monitoring solution could analyze activities performed in daily life, such as drinking, eating and sleeping habits.

@@

Capacitive proximity sensors on the other hand can indirectly measure the distance and nature of a grounded object within reach. This means that the measurement result depends on the object's distance, its size and the material it is made of. In this work, we show that an accelerometer and a capacitive proximity sensor can be used to improve activity recognition in activities of daily living.
We obtained an open-hardware and open-source wrist-worn activity data logger \cite{hedgehog} and integrated a capacitive proximity sensor invisibly into the wristband.

@@

@@

Activity recognition research relying on wearable sensors mostly considers inertial data from the participants body to infer performed activities, such as in the works of \cite{Ravi2005,Bao2004,Srinivasan2010,Amft2005}. The acceleration data is often augmented with data from sensors such as gyroscopes \cite{Holleczek_2010}, magnetometers \cite{Altun_2010}, ambient light \cite{Borazio2012} or ambient and skin temperature \cite{Krause_2003}, aiming to extract a more detailed environmental user context. In \cite{wyss2010recognition}, the authors use heart rate information as an addition to the accelerometer data to detect activities like lifting and lowering loads or even digging. In \cite{Ward_2006}, workshop assembly activities are detected by augmenting acceleration sensors with microphones.

The works of Fishkin et al. \cite{Fishkin_2005} and Patterson et al. \cite{patterson2005fga} show that detecting touched and used objects can be very helpful for activity recognition. By using RFID readers that are embedded in gloves or bracelets at the wrist and RFID tags attached to various objects of interest, one can detect the object grasped and used by the user, thus aiding the activity recognition in various application scenarios, such as activities of daily living \cite{phealth:maja} and \cite{Philipose_2004}, activity tracking in car manufacturing \cite{Stiefmeier08}, or household and gardening activities \cite{berlin_laerhoven_tei_2010}.

@@

@@

This section presents the two components of our hardware prototype, the wrist-worn activity data logger tailored to capture acceleration data, and the capacitive proximity sensor used for distance measurements.

@@

%  (PIC18F46J50) (ADXL345)

The sensor node itself is built around the low-power Microchip microcontroller featuring an accelerometer sensor to capture human motion, light and ambient temperature sensors and a microSD flash card for locally storing the sensor data. The sensor is powered by a $200mAh$ lithium polymer battery, which allows for two weeks of continuous recording on a single battery charge. A USB port is used to configure the sensor (e.g. setting the sensitivity of the accelerometer), to access the stored sensor data, and to recharge the battery. A plastic case nicely packages and protects the sensor to be worn at the wrist (Figure \ref{fig:sensornode}).

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{Images/hardware_sensor_2.jpg}
	\caption{The inertial data logger featuring a low-power microcontroller, a 3 axis accelerometer, a microSD flash card for storing the sensor data and a USB connector for accessing the data (on the right) is powered by a small lithium polymer battery and is packaged into a plastic case to be worn at the wrist (a version with a OLED display).}
	\label{fig:sensornode}
\end{figure}

The 3D accelerometer sensor is being sampled at 100Hz, resulting in 10ms equidistant measurements. For efficiency reasons, the sensor data is run-length encoded before being stored locally to the microSD card. 
The HedgeHog can be extended with further sensors tailoring different application scenarios. For our scenario, we have added a capacitive proximity sensor that is described in detail in the next section.

@@

The capacitive proximity sensor performs measurements in loading mode. Two electrodes are integrated into the wristband, one sensing electrode and one shielding electrode. The sensor draws a supply current of 1mA at 3.3V when active and qualifies it for wearable proximity sensing applications. In the following a virtual capacitor denotes the capacitance between the sensing electrode and the environment. The capacitance of the sensing electrode to environmental objects increases with closer distances.
% The capacitive proximity sensor performs measurements in loading mode. Two electrodes are integrated into the wristband, one sensing electrode and one shielding electrode that aims to eliminate the influence of the underlying skin. Without shielding no meaningful measurements could be performed, since the arm's skin is a very proximate grounded object to the sensing electrode. The sensor draws a supply current of 1mA at 3.3V when active and qualifies it for wearable proximity sensing applications. In the following a virtual capacitor denotes the capacitance between the sensing electrode and the environment. The capacitance of the sensing electrode to environmental objects increases with closer distances.

@@

\begin{figure}[t]
	\centering
		\includegraphics[width=\textwidth]{Images/capacitive_sensor_wristband_2.jpg}
	\caption{The hardware prototype at a glance: HedgeHog activity logger at the lower right, the capacitive sensor unit at the lower left, and the wristband with the sensing and the shield electrodes on-top each other covered with adhesive tape for isolation purposes.}
	\label{fig:cap_sensor}
\end{figure}

@@

The opening door activity has very poor recognition rates without the data from the proximity sensor. The average F-measure could be increased from 35.5\% to 46.2\%. A plot of the activity is given in Figure \ref{fig:opendoor}. The capacitive proximity sensor shows two approaches to the door knob, one for opening the door (2s) and one for closing the door (9s). The acceleration sensor captures relevant data in the time in which the person moves into the room and the hand changes from the outer to the inner door knob (5 - 7s). The recorded data for this activity also shows strong correlations between all experiment participants. The confusion matrices show that the ``open door'' activity was often confused with the ``sitting'' activity, probably because of the amount of motion on the one hand and the proximity to nearby objects (door, couch or cushions) on the other hand. By using the capacitive sensor data, the recall for that class and confusion with the sitting activity could be improved.

@@

After closing the door, the participants were supposed to sit down on the couch. It turned out that there are great variations of the sitting posture and the corresponding hand positions. Many users tapped with their fingers or hands while sitting, changed their sitting positions very frequently, or were even talking and gesticulating, as shown in Figure \ref{fig:sitting}. In this case, it is obvious that the data from the acceleration sensor is very difficult to interpret as there are numerous changes in the axial orientation of the sensor. However, the capacitive proximity sensor is able to indicate when a hand is placed on the surface of the couch. Especially for this particular participant, the F-measure increased from 50.3\% to 60.4\%, while the average F-measure improved from 68.0\% to 74.4\%.

In the following, the participant were instructed to lie  down on the couch. Again, there were great variations in how this activity was performed by the participants. For example, some of them crossed their hands under their head, or placed them on their body. For this class, the average F-measure could only be increased by 2.4\%, from 81.2 to 83.6\%. Considering some participants, the activity was often confused with the ``making bread'' class. By using the proximity modality, the confusion between the two classes could be reduced.

@@

After that, the participants were asked to walk over to the kitchen and to put food and dishes from a shelf and a locker on the table. This activity involved direct interactions with various objects as well as proximity to furniture in the room (see Table \ref{tab:activities}). The capacitive proximity sensor was able to capture the proximity to the shelf and to the table. The average F-measure for this activity is rather low, but improved by 6.8\% from 53.8\% to 60.6\%. 
The worst performing participants for this activity reached an F-measure of 46.8\% without and 53.5\% with the capacitive sensor, while the best performing one reached 62.0\% and 64.5\% respectively. The quite low performance figures result from confusions with other activities, with a higher tendency to the ``prepare bread'' activity across all participants. This is most likely due to various objects involved in both activities. The capacitive sensor modality has a much more positive impact reducing the confusion with other activities.

\begin{figure}
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_5.pdf}
	\caption{An example of the ``preparing bread'' activity, where the participants had to put marmalade on a slice of bread. The proximity sensor indicates the closeness to the table, while the acceleration sensor shows recurring hand motions.}
	\label{fig:prep_bread}
\end{figure}

Figure \ref{fig:prep_bread} shows an example instance of preparing a bread with marmalade. It is notable that the acceleration data does not seem to provide any characteristic patterns, while the proximity sensor indicates a table, plate, or other objects in immediate distance. This activity showed a high improvement in the average F-measure by 10\%, from 49.0\% to 59.8\%, where the data delivered by the capacitive proximity sensor is taken into account. The making bread class was often confused with the sitting class for some users, probably due to lots of motion during the sitting, as mentioned previously. For other users, making bread was confused with eating or drinking. Using the new input modality, confusion across users could be reduced.

\begin{figure}
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_6.pdf}
	\caption{An example of a participant eating a slice of bread with marmalade, taking 5 bites from it. After each bite, the hand is placed on the table, which can be recognized both in the acceleration as well as the proximity data plots.}
	\label{fig:eating}
\end{figure}

When considering the eating activity, the impact of the new capacitive proximity sensor on the classification performance is quite low, as the chosen features are able to distinguish it from other activities. Some of the participants ate their bread leaving their hand close to the mouth, while others moved their hand up and down putting their bread aside on the plate, which also results in the performance range from 71.2 to 88.1\% without and 77.5 to 90.6\% with the proximity data. A good example of an eating activity is shown in Figure \ref{fig:eating}, where the participant took a few bites from the bread while putting it down every time. The average F-measure for that activity increased slightly by 2.7\% (from 79.5\% to 82.2\%).

\begin{figure}[t]
	\centering
		\includegraphics[width=\textwidth]{../Auswertung/images/eugen_7.pdf}
	\caption{An example of the ``drinking'' activity. The participant first pours some water into the glass and then takes three drinks of water. After each sip, he returns his arm to the table which can be observed in the characteristic patterns of the proximity sensor.}
	\label{fig:drinking}
\end{figure}

@@

Regarding the walking activity one can identify periodic changes in the acceleration as well as in the measured capacitance, illustrated in Figure \ref{fig:walking}. While walking, the capacitance between the wristband and the leg increases when the wristband is located close to the body and decrease when the wristband moves away. There were problems in distinguishing this activity from get things that could be bettered by using the new input modality. The classification improvement for this activity accounts to 12.2\% boosting the average F-measure from 41.7\% without to 53.9\% with the new sensor. Due to the low performance results and the characteristic periodic signal shape, it would help to consider frequency domain features, as it is often applied in related work.

\begin{figure}[t]
	\centering
		\includegraphics[width=1.00\textwidth]{../Auswertung/images/eugen_9.pdf}
	\caption{During the sleeping activity the data from both sensors remains constant for large time spans. One can draw conclusions about the coverage of the arm with either cushions, blankets or the proximity to the mattress, the head or body of the participant.}
	\label{fig:sleeping}
\end{figure}

The sleeping activity (an example shown in Figure \ref{fig:sleeping}) was classified with an average F-measure of 83.2\%, which increased to 86.9\% when using the proximity data. In this case, the capacitive proximity sensor is able to capture the surrounding cushions and blankets, as well as the body or head of the participant. The accelerometer data and the proximity data has larger periods with a constant signal, which is a cause for confusion with the ``lying on the couch'' activity.

\begin{figure}
	\setlength{\tabcolsep}{2pt}
	\centering
	\begin{scriptsize}
	\begin{tabular}{|ccccccccc|p{0.01cm}|ccccccccc|p{0.01cm}|l}
% 		\toprule
		\multicolumn{9}{c}{without proximity data} &\multicolumn{1}{c}{}& 
			\multicolumn{9}{c}{with proximity data} &\multicolumn{1}{c}{}&  \\ \cmidrule{1-9} \cmidrule{11-19}

		\multicolumn{1}{c}{a} & b & c & d & e & f & g & h & \multicolumn{1}{c}{i} &\multicolumn{1}{c}{}& 
			\multicolumn{1}{c}{a} & b & c & d & e & f & g & h & \multicolumn{1}{c}{i} 	&\multicolumn{1}{c}{}& $\leftarrow$ classified as\\
		\cline{1-9} \cline{11-19} \cline{21-21}
\textbf{8} & 0 & 0 & 0 & 0 & 0 & 0 & 2 & 1		&& \textbf{11} & 0 & 0 & 2 & 1 & 0 & 0 & 0 & 1 	&& a = open door	\\ 
0 & \textbf{140} & 0 & 5 & 2 & 0 & 0 & 0 & 0  	&& 0 & \textbf{140} & 0 & 5 & 2 & 0 & 0 & 0 & 0 && b = sitting		\\ 
0 & 0 & \textbf{294} & 1 & 9 & 10 & 1 & 1 & 0 	&& 0 & 0 & \textbf{309} & 1 & 0 & 4 & 0 & 0 & 1 && c = lying		\\ 
1 & 7 & 1 & \textbf{50} & 17 & 9 & 1 & 4 & 1	&& 2 & 6 & 0 & \textbf{67} & 14 & 9 & 1 & 2 & 1 && d = get things	\\ 
0 & 2 & 7 & 7 & \textbf{102} & 42 & 5 & 0 & 2	&& 1 & 1 & 0 & 13 & \textbf{120} & 24 & 7 & 1 & 1&& e = make bread	\\ 
0 & 0 & 3 & 5 & 21 & \textbf{288} & 4 & 0 & 0 	&& 0 & 0 & 2 & 1 & 19 & \textbf{300} & 3 & 0 & 0 && f = eating		\\ 
0 & 5 & 11 & 1 & 29 & 40 & \textbf{18} & 0 & 0	&& 0 & 4 & 4 & 0 & 26 & 57 & \textbf{16} & 0 & 0 && g = drinking	\\
2 & 0 & 0 & 2 & 0 & 0 & 0 & \textbf{11} & 0	 	&& 2 & 0 & 0 & 0 & 0 & 0 & 0 & \textbf{14} & 0 	 && h = walking		\\
0 & 2 & 2 & 0 & 2 & 0 & 0 & 0 & \textbf{271}	&& 0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 & \textbf{276}  && i = sleeping	\\ 
		\cline{1-9} \cline{11-19} 
% 		\bottomrule
	\end{tabular}
	\end{scriptsize}

	\caption{Activity recognition evaluation revealing the positive impact of the capacitive proximity sensor. Here, we are comparing SVM classification presented as confusion matrices for an exemplary user, without the proximity data on the left, and with the proximity data on the right. Note that the reject class (background data not annotated as an activity) is not included in the confusion matrix.}
	\label{fig:ConfusionMatrixForAnExemplaryUser}
\end{figure}

Figure \ref{fig:ConfusionMatrixForAnExemplaryUser}  depicts two confusion matrices for an exemplary participant from our evaluation, once without and once with including the proximity data. In most activities, an enhancement in the number of correctly classified instances is observable. The ``lying'' activity's recognition performance could benefit a lot from the proximity data, improving both precision and recall.

A better classification performance can also be observed for the ``get things'' class that includes interactions with a multitude of objects in the environment. Regarding the activity ``make bread'', the capacitive proximity could reduce the number of confusions with the ``eating'' class significantly.

Due to the high similarity of eating and drinking (cf. Figure \ref{fig:eating} and \ref{fig:drinking}), the number of confusions between those two classes increases when considering the proximity data. However, for the ``drinking'' class the new modality limits the confusions to related activities such as ``eating'' and ``make bread'' only, while lowering the number of false recognitions for the other activities.

@@

Our paper presented a wrist-worn activity data logger prototype, which consists of an accelerometer in combination with a capacitive proximity sensor integrated into the wristband. Our experiments with seven participants and nine basic daily activities show that this additional input modality can significantly boost the activity recognition performance.
Regarding the classification performance, we obtained an improvement in the average F-measure of 6.3\%, from 67.2 to 73.5\%. Specifically, the activity classes ``walking'', ``preparing bread'' and ``open door'' could benefit a lot from proximity-related sensor data. For such classes, the classification performance could be boosted by 12.2, 9.0 and 10.8\% respectively.

With this proof of concept we show that the proximity information can provide an information gain regarding the evaluated activities. In future work we aim at evaluating other relevant feature types, such as frequency domain features, as well as feature sets to extract the most discriminative ones. Additionally, using other classifiers (such as HMMs) or more fitting classifier configurations might also improve activity recognition performance.
